---
title: "Memory Efficient Epigenomics And Grob Visualizations"
author: "Kathryn Lande"
date: "2026-05-29"
output: 
  rmarkdown::github_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

# Setup

Install chromDict functions from github:
```{r, message=FALSE, warning=FALSE}
# remotes::install_github("katlande/chromDicts")
```

load packages:
```{r, message=FALSE, warning=FALSE}
library(chromDicts)
library(PCBS)
library(ggplot2)
library(dplyr)
library(data.table)
library(ggpubr)
```

Import test data:
```{r}
# get the dictlist and gtf for out testing data
data(testing)
```


# Customizable Peak Tracks
Using chromDicts, we can quickly and easily extract genomic loci and plot them as highly customizable grobs. Below are some examples of how to make peak tracks directly in ggplot.

## Peak Tracks at Genes

Here we show a minimal example:
```{r, message=FALSE}

makeGeneTrack(DictList = testing_dicts, # list of chromDicts to plot
              gtf = GTF,
              gene="Npas2") # gene to plot over

```


We can also use more advanced options to customize the plot, including colouring the peak tracks by a specified variable using a custom palette, annotating specific sites, and changing the peak track height.

```{r, message=FALSE}

meta <- data.frame(names(testing_dicts), c("CTL", "TRT"))

annots <- data.frame(chr=c("chr1", "chr1"),
                     s=c(39250000, 39272000),
                     e=c(39265000, 39290000))

makeGeneTrack(DictList=testing_dicts, 
              gtf=GTF, 
              gene="Npas2", 
              rel=3, # relative height of peak track window to annotation window
              meta = meta, # meta data for colouring tracks
              cols=c("grey", "skyblue"), # custom colour palette
              annotation_bed = annots) # bed-formatted sites to annotate

```


## Peak Tracks at Specific Loci:

In some cases, you may want to look at an intergenic site, a subset of a gene, or another region that can't be well captured with makeGeneTrack(). You can also make peak tracks using specified loci:
```{r, message=FALSE, warning=FALSE}

# settings are basically the same, but instead of using the gene name, supply a specific locus. 
# Here we look at the beginning of Gbgt1:
makeLocusTrack(DictList=testing_dicts, 
              gtf=GTF, 
              chr = "chr2",
              start = 28496000,
              end = 28499000,
              rel=3, # relative height of peak track window to annotation window
              meta = meta, # meta data for colouring tracks
              cols=c("grey", "skyblue"),
              annotation_bed = data.frame("chr2", 28497650, 28498000)) # bed-formatted sites to annotate

```

# Metagenes

Metagenes are a useful tool for looking at average genomic trends across a large set of connected regions. In this example, we'll pull +/- 500bp of all start codons in our test GTF to compare our treat and control samples.

```{r, message=FALSE, warning=FALSE}

# extract loci to plot into a bed-formated data.frame:
start_codons <- subset(GTF, V3 == "start_codon")
loci <- data.frame(chr=paste0("chr", start_codons$V1), 
                   start=start_codons$V4-500, 
                   end=start_codons$V5+500)

# Use the metagene functionality from PCBS to plot a minimal metagene for the control sample:
valName <- colnames(testing_dicts$control[[1]])[3]

methylDiff_metagene(testing_dicts$control, # input one chromDict
                    loci, # bed-formatted loci to query
                    return.data = F, # if FALSE, returns a plot (for simple metagenes)
                    yaxis = "Mean BPM",
                    title = "Simple Metagene",
                    value = valName) # name of value column in chromDict

```

We can also make more advanced metagenes comparing different conditions or area sets, as below:

```{r, message=FALSE, warning=FALSE}

# extract metagene information from each sample:
metagene_ctl <- methylDiff_metagene(testing_dicts$control, loci, return.data = T, value = valName)
metagene_trt <- methylDiff_metagene(testing_dicts$treatment, loci, return.data = T, value = valName)

# Overlay control and treatment start codon metagenes in one plot:
multiple_metagenes(data_list = list(metagene_ctl, metagene_trt), # list of raw data
                   set_names = c("Control", "Treatment"), # names for elements of the data_list list
                   col=c(Control="skyblue", Treatment="salmon"), # colours for metagenes
                   title="Start Codons, +/-500",
                   legend.title = F, 
                   yaxis = "Mean BPM")


```

Above, we see start codons have a distinctly higher signal in treatment samples than they do in control samples. We can also check if this extends to other features, for example, exons:


```{r, message=FALSE, warning=FALSE}

# extract all exons:
exons <- subset(GTF, V3 == "exon")
exon_loci <- data.frame(chr=paste0("chr", exons$V1), 
                        start=exons$V4, 
                        end=exons$V5)

metagene_ctl_exon <- methylDiff_metagene(testing_dicts$control, exon_loci, return.data = T, value = valName)
metagene_trt_exon <- methylDiff_metagene(testing_dicts$treatment, exon_loci, return.data = T, value = valName)

# plot both exon and start codon metagenes together:
multiple_metagenes(data_list = list(metagene_ctl, metagene_ctl_exon, metagene_trt, metagene_trt_exon),
                   set_names = c("Ctl - Start Codons", "Ctl - Exons", "Trt - Start Codons", "Trt - Exons"),
                   col=c("skyblue", "navy", "salmon", "darkred"), # colours for metagenes
                   title="Start Codons & Exons",
                   legend.title = F, 
                   yaxis = "Mean BPM")


```
Above, we see that start codons generally have a higher signature than exons, and the difference between treatment and control is also bigger at exons. 


# Quantifying Epigenomic Signatures Across Regions

In some cases, you may want to quantify the scale and significances of these differences across regions. For example, in the above metagenes, we see a difference in our signature at start codons between conditions. But is this a significant difference? We can pull the average signatures across these regions and plot them:

```{r, message=FALSE, warning=FALSE}

# plot the mean across all loci
RegionAverages(testing_dicts, loci, cols = c("skyblue", "salmon"))+
  stat_compare_means(label.y = 0.15) # add a significance metric

```

Above, we see that this difference between treatment and control is *not* significant with a p=0.2, but merely biased by a few outlier loci in the treatment that have particularly high accumulation of our signal.















